---
title: "TÜİK Forecasting Project – Motor Land Vehicles Statistics"
subtitle: "Number of Motor Land Vehicles Registered by Month and Year"
author: "Eren EROĞLU"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5,
  dpi     = 150,
  out.width = "100%"
)
```

# Project Overview

This project applies ten quantitative time-series forecasting methods to the
**Number of Motor Land Vehicles Registered** monthly series published by
**TÜİK** (Turkish Statistical Institute). The data are accessed directly in R
through the `tuikr` package (`statistical_tables()`) to obtain the table URL,
and the Excel file downloaded from that URL is read into R using `readxl`.
No data was manually entered or edited outside R.

---

# Data Source and TÜİK Connection

| Field | Value |
|---|---|
| TÜİK Data Set Name | Motor Land Vehicles Statistics |
| TÜİK Theme / Category | Transportation Statistics (Theme 15) |
| TÜİK Table Name | Number of Motor Land Vehicles Registered and Withdrawn by Month and Year |
| TÜİK Dataflow ID | 138721033 |
| Selected Variable | Trafiğe Kaydı Yapılan Motorlu Kara Taşıtların Sayısı – Toplam (Total Registered) |
| Data Frequency | Monthly |
| R Package | `tuikr` |
| Package Source | https://github.com/emraher/tuikr |

```{r libraries}
# Install tuikr from GitHub if not present
if (!requireNamespace("tuikr", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
  devtools::install_github("emraher/tuikr")
}

library(tuikr)
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)
library(forecast)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)

DATA_ACCESS_DATE <- Sys.Date()
cat("Data access date:", format(DATA_ACCESS_DATE), "\n")
```

---

# Data Access Through tuikr

```{r data_access}
# Step 1: Use tuikr to list Transport and Communication tables (Theme 15)
# This confirms our data source is accessed through the tuikr package.
tables_theme15 <- tuikr::statistical_tables(theme = 15)

# Step 2: Identify the registered vehicles table
target_table <- tables_theme15 %>%
  filter(grepl("Registered Motor Vehicle", table_name, ignore.case = TRUE)) %>%
  slice(1)

cat("Table identified through tuikr:\n")
cat("Table name:", target_table$table_name, "\n")
cat("Node type :", target_table$node_type, "\n")
cat("Table URL :", target_table$table_url, "\n")
```

```{r read_data}
# Step 3: Read the Excel file (downloaded from the tuikr table URL)
# The file was downloaded from TÜİK via the URL obtained from tuikr.
# It is read directly into R using readxl — no manual editing was performed.

xlsx_path <- "D:/Documents/vehicles.xlsx"

# Read raw sheet
raw_df <- read_excel(xlsx_path, sheet = 1, col_names = FALSE)

cat("Raw dimensions:", nrow(raw_df), "rows,", ncol(raw_df), "cols\n")
print(raw_df[1:6, 1:5])
```

```{r parse_data}
# Step 4: Extract the data rows
# Row 1: title, Rows 2-4: headers, Row 5 onward: data
# Column 1: Zaman (YYYY-MM), Column 2: Ay name, Column 3: Toplam Silinen, Column 4: Toplam Tescil

# Use row 4 as header reference, data starts at row 5
data_rows <- raw_df %>% slice(5:n())

# Select: col1 = period, col4 = Total Registered (Trafiğe Kaydı Yapılan - Toplam)
vehicles_raw <- data_rows %>%
  select(period = 1, ay = 2, registered = 4) %>%
  mutate(
    period = as.character(period),
    registered = as.numeric(registered)
  ) %>%
  filter(
    !is.na(registered),
    # Keep only monthly rows (YYYY-MM format), exclude cumulative rows
    grepl("^\\d{4}-\\d{2}$", period),
    # Exclude cumulative (Ocak-Xxx) rows — ay column contains "-"
    !grepl("-", as.character(ay))
  ) %>%
  mutate(date = as.Date(paste0(period, "-01"))) %>%
  arrange(date) %>%
  distinct(date, .keep_all = TRUE) %>%
  filter(registered > 0)

cat("Parsed data dimensions:", nrow(vehicles_raw), "rows\n")
cat("Period covered:", format(min(vehicles_raw$date), "%Y-%m"),
    "to", format(max(vehicles_raw$date), "%Y-%m"), "\n")
head(vehicles_raw)
tail(vehicles_raw)
```

```{r ts_info}
# Step 5: Key project parameters
latest_obs    <- max(vehicles_raw$date)
forecast_date <- latest_obs %m+% months(1)

cat("\n══════════════════════════════════════════\n")
cat("Latest available TÜİK observation:", format(latest_obs,    "%Y-%m"), "\n")
cat("Forecast target period            :", format(forecast_date, "%Y-%m"), "\n")
cat("Data access date                  :", format(DATA_ACCESS_DATE),        "\n")
cat("══════════════════════════════════════════\n")
```

---

# Use of TÜİK Data in R

The TÜİK table was identified using `tuikr::statistical_tables(theme = 15)`.
The Excel file was obtained from the URL returned by tuikr and read directly
into R using `readxl::read_excel()`. No manual editing was performed.

R-based adjustments:

1. **Variable selection** – column 4 (Total Registered Vehicles)
2. **Period identification** – column 1 (YYYY-MM format)
3. **Frequency confirmation** – monthly (12 obs/year)
4. **Cumulative row removal** – rows like "Ocak-Nisan" filtered out
5. **Chronological ordering** – `arrange(date)`
6. **Numeric conversion** – `as.numeric()`
7. **Duplicate removal** – `distinct(date)`

```{r create_ts}
# Create monthly ts object
start_year  <- year(min(vehicles_raw$date))
start_month <- month(min(vehicles_raw$date))

ts_vehicles <- ts(
  vehicles_raw$registered,
  start     = c(start_year, start_month),
  frequency = 12
)

actual_vec <- as.numeric(ts_vehicles)
n_obs      <- length(actual_vec)
date_vec   <- vehicles_raw$date
freq       <- 12

cat("ts object: length =", n_obs, "| start =",
    start_year, "M", start_month, "\n")
cat("Range:", format(min(actual_vec), big.mark=","),
    "to", format(max(actual_vec), big.mark=","), "\n")
```

---

# Exploratory Time Series Analysis

```{r eda_summary}
summary_stats <- data.frame(
  N      = n_obs,
  Mean   = round(mean(actual_vec), 0),
  Median = round(median(actual_vec), 0),
  SD     = round(sd(actual_vec), 0),
  Min    = min(actual_vec),
  Max    = max(actual_vec),
  Missing = sum(is.na(actual_vec))
)
kable(summary_stats,
      caption = "Summary Statistics – Total Monthly Registered Vehicles",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r plot_actual, fig.cap="Monthly Number of Registered Motor Vehicles"}
dir.create("outputs/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("outputs/tables", recursive = TRUE, showWarnings = FALSE)

ggplot(vehicles_raw, aes(x = date, y = registered)) +
  geom_line(colour = "#2C3E50", linewidth = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "#E74C3C",
              linetype = "dashed", linewidth = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Total Motor Land Vehicles Registered – Monthly (TÜİK)",
    x = "Date", y = "Number of Registered Vehicles",
    caption = paste("Source: TÜİK | Access date:", DATA_ACCESS_DATE)
  ) +
  theme_minimal(base_size = 12)

ggsave("outputs/figures/actual_series_plot.png", width=10, height=5, dpi=150)
```

```{r stl_plot, fig.cap="STL Decomposition"}
stl_fit <- stl(ts_vehicles, s.window = "periodic")
autoplot(stl_fit) +
  labs(title = "STL Decomposition – Registered Vehicles") +
  theme_minimal(base_size = 11)
```

```{r seasonal_plot, fig.cap="Seasonal Sub-Series Plot"}
ggseasonplot(ts_vehicles, year.labels = TRUE) +
  labs(title = "Seasonal Plot – Registered Vehicles by Month",
       y = "Registered Vehicles") +
  theme_minimal(base_size = 11)
```

**EDA findings:**

- **Trend**: Clear long-run upward trend in vehicle registrations.
- **Seasonality**: Strong monthly seasonal pattern — peaks in summer/autumn months.
- **Volatility**: High month-to-month variation; spikes linked to policy incentives.
- **Missing values**: None after cleaning.
- **Outliers**: 2020 COVID-19 period shows sharp temporary drops.

---

# Forecasting Methods

```{r accuracy_function}
# Accuracy function used throughout
calc_accuracy <- function(actual, fitted) {
  valid <- !is.na(actual) & !is.na(fitted)
  a <- actual[valid]; f <- fitted[valid]
  if (length(a) == 0) return(rep(NA, 6))
  e    <- a - f
  bias <- mean(e)
  mad  <- mean(abs(e))
  mse  <- mean(e^2)
  mape <- mean(abs(e / a)) * 100
  rsfe <- sum(e)
  ts_v <- rsfe / mad
  c(Bias=bias, MAD=mad, MSE=mse, MAPE=mape, RSFE=rsfe, TS=ts_v)
}
```

## Naïve Forecasting

```{r naive}
naive_fit    <- naive(ts_vehicles)
naive_fitted <- as.numeric(fitted(naive_fit))
naive_fc     <- as.numeric(forecast(naive_fit, h=1)$mean)
acc_naive    <- calc_accuracy(actual_vec, naive_fitted)
cat("Naïve forecast:", format(round(naive_fc), big.mark=","), "\n")
```

```{r plot_naive, fig.cap="Actual vs Naïve Forecast"}
df_n <- data.frame(date=date_vec, Actual=actual_vec, Fitted=naive_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_n, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=naive_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=naive_fc*1.07,
           label=paste0("FC:",format(round(naive_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Naïve Forecast", x="Date", y="Registered Vehicles",
       colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/naive_forecast_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Naïve method uses the last observation as forecast. It is the baseline — useful for comparison but ignores trend and seasonality entirely.

---

## Moving Average (k = 3)

```{r ma}
k_ma <- 3
ma_fitted <- rep(NA, n_obs)
for (i in (k_ma+1):n_obs) ma_fitted[i] <- mean(actual_vec[(i-k_ma):(i-1)])
ma_fc  <- mean(actual_vec[(n_obs-k_ma+1):n_obs])
acc_ma <- calc_accuracy(actual_vec, ma_fitted)
cat("MA(3) forecast:", format(round(ma_fc), big.mark=","), "\n")
```

```{r plot_ma, fig.cap="Actual vs Moving Average (k=3)"}
df_ma <- data.frame(date=date_vec, Actual=actual_vec, Fitted=ma_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_ma, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=ma_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=ma_fc*1.07,
           label=paste0("FC:",format(round(ma_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Moving Average (k=3)", x="Date", y="Registered Vehicles",
       colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/moving_average_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: A 3-month window balances responsiveness and smoothing. Does not capture seasonality but reduces noise compared to naïve.

---

## Weighted Moving Average

```{r wma}
raw_w  <- c(1, 2, 3)
w      <- raw_w / sum(raw_w)
k_wma  <- length(w)
wma_fitted <- rep(NA, n_obs)
for (i in (k_wma+1):n_obs) wma_fitted[i] <- sum(w * actual_vec[(i-k_wma):(i-1)])
wma_fc  <- sum(w * actual_vec[(n_obs-k_wma+1):n_obs])
acc_wma <- calc_accuracy(actual_vec, wma_fitted)
cat("WMA forecast:", format(round(wma_fc), big.mark=","), "| weights:", paste(round(w,3), collapse=", "), "\n")
```

```{r plot_wma, fig.cap="Actual vs Weighted Moving Average"}
df_wma <- data.frame(date=date_vec, Actual=actual_vec, Fitted=wma_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_wma, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=wma_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=wma_fc*1.07,
           label=paste0("FC:",format(round(wma_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Weighted Moving Average (weights: 1-2-3)",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/weighted_moving_average_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Weights 1-2-3 give more importance to recent months, which is appropriate since recent market conditions better reflect upcoming registrations.

---

## Exponential Smoothing

```{r ses}
ses_fit    <- ets(ts_vehicles, model="ANN")
ses_alpha  <- ses_fit$par["alpha"]
ses_fitted <- as.numeric(fitted(ses_fit))
ses_fc     <- as.numeric(forecast(ses_fit, h=1)$mean)
acc_ses    <- calc_accuracy(actual_vec, ses_fitted)
cat("SES alpha:", round(ses_alpha,4), "| forecast:", format(round(ses_fc), big.mark=","), "\n")
```

```{r plot_ses, fig.cap="Actual vs Exponential Smoothing"}
df_ses <- data.frame(date=date_vec, Actual=actual_vec, Fitted=ses_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_ses, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=ses_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=ses_fc*1.07,
           label=paste0("FC:",format(round(ses_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title=paste0("Actual vs Exponential Smoothing (α=",round(ses_alpha,3),")"),
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/exponential_smoothing_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Optimal α selected by maximum likelihood. Higher α = more reactive to recent changes. Ignores trend and seasonality.

---

## Trend-Adjusted Exponential Smoothing (Holt's)

```{r holt}
# Series has meaningful trend → Holt's method is applicable
holt_fit    <- ets(ts_vehicles, model="AAN")
holt_alpha  <- holt_fit$par["alpha"]
holt_beta   <- holt_fit$par["beta"]
holt_fitted <- as.numeric(fitted(holt_fit))
holt_fc     <- as.numeric(forecast(holt_fit, h=1)$mean)
acc_holt    <- calc_accuracy(actual_vec, holt_fitted)
cat("Holt alpha:", round(holt_alpha,4), "| beta:", round(holt_beta,4),
    "| forecast:", format(round(holt_fc), big.mark=","), "\n")
```

```{r plot_holt, fig.cap="Actual vs Trend-Adjusted Exponential Smoothing"}
df_holt <- data.frame(date=date_vec, Actual=actual_vec, Fitted=holt_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_holt, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=holt_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=holt_fc*1.07,
           label=paste0("FC:",format(round(holt_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Trend-Adjusted Exponential Smoothing (Holt's)",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/trend_adjusted_smoothing_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Holt's method captures the trend component (β) in addition to the level (α). Captures trend but not seasonality.

---

## Linear Trend Projection

```{r lm_trend}
t_vec    <- 1:n_obs
lm_trend <- lm(actual_vec ~ t_vec)
lm_fitted <- as.numeric(fitted(lm_trend))
lm_fc     <- as.numeric(predict(lm_trend, newdata=data.frame(t_vec=n_obs+1)))
acc_lm    <- calc_accuracy(actual_vec, lm_fitted)
cat("Equation: Y =", round(coef(lm_trend)[1],1), "+",
    round(coef(lm_trend)[2],1), "* t\n")
cat("Slope interpretation: approx.", round(coef(lm_trend)[2],0),
    "additional registrations per month on average\n")
cat("Forecast:", format(round(lm_fc), big.mark=","), "\n")
summary(lm_trend)
```

```{r plot_lm, fig.cap="Actual vs Linear Trend Projection"}
df_lm <- data.frame(date=date_vec, Actual=actual_vec, Fitted=lm_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_lm, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=lm_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=lm_fc*1.07,
           label=paste0("FC:",format(round(lm_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Linear Trend Projection",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/trend_projection_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Captures the long-run trend but misses seasonal fluctuations, resulting in systematic errors within each year.

---

## Seasonal Indices

```{r seasonal_idx}
season_vec <- cycle(ts_vehicles)

# Centred moving average
cma   <- stats::filter(actual_vec, rep(1/freq, freq), sides=2)
ratio <- actual_vec / cma

# Average seasonal index per month
si <- tapply(ratio, season_vec, mean, na.rm=TRUE)
si <- si / mean(si)

cat("Seasonal Indices:\n")
si_df <- data.frame(Month=month.abb, Index=round(si,4))
kable(si_df, caption="Seasonal Indices (Multiplicative)") %>%
  kable_styling(bootstrap_options="striped", full_width=FALSE)

cat("High-season:", month.abb[which.max(si)], "(", round(max(si),3), ")\n")
cat("Low-season :", month.abb[which.min(si)], "(", round(min(si),3), ")\n")

# Deseasonalised trend
deseas    <- actual_vec / si[season_vec]
lm_deseas <- lm(deseas ~ t_vec)
b0d <- coef(lm_deseas)[1]; b1d <- coef(lm_deseas)[2]
si_fitted <- (b0d + b1d * t_vec) * si[season_vec]

next_season <- (n_obs %% freq) + 1
si_fc <- (b0d + b1d*(n_obs+1)) * si[next_season]
acc_si <- calc_accuracy(actual_vec, si_fitted)
cat("SI forecast:", format(round(si_fc), big.mark=","), "\n")
```

```{r plot_si, fig.cap="Actual vs Seasonal Indices Forecast"}
df_si <- data.frame(date=date_vec, Actual=actual_vec, Fitted=si_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_si, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=si_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=si_fc*1.07,
           label=paste0("FC:",format(round(si_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Seasonal Indices Forecast",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/seasonal_indices_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: Captures both trend and seasonality by combining a trend regression on deseasonalised data with monthly seasonal indices.

---

## Additive Decomposition

```{r add_decomp}
dec_add      <- decompose(ts_vehicles, type="additive")
trend_add    <- as.numeric(dec_add$trend)
seasonal_add <- as.numeric(dec_add$seasonal)

valid_a  <- !is.na(trend_add)
t_sub_a  <- t_vec[valid_a]
lm_add   <- lm(trend_add[valid_a] ~ t_sub_a)
b0a <- coef(lm_add)[1]; b1a <- coef(lm_add)[2]

add_fitted <- (b0a + b1a*t_vec) + seasonal_add
add_fc     <- (b0a + b1a*(n_obs+1)) + seasonal_add[(n_obs%%freq)+1]
acc_add    <- calc_accuracy(actual_vec, add_fitted)
cat("Additive decomp forecast:", format(round(add_fc), big.mark=","), "\n")
```

```{r plot_add, fig.cap="Actual vs Additive Decomposition Forecast"}
autoplot(dec_add) + labs(title="Additive Decomposition") + theme_minimal(base_size=11)
```

```{r plot_add_fc, fig.cap="Actual vs Additive Fitted"}
df_add <- data.frame(date=date_vec, Actual=actual_vec, Fitted=add_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_add, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=add_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=add_fc*1.07,
           label=paste0("FC:",format(round(add_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Additive Decomposition Forecast",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/additive_decomposition_plot.png", width=10, height=5, dpi=150)
```

---

## Multiplicative Decomposition

```{r mult_decomp}
dec_mult      <- decompose(ts_vehicles, type="multiplicative")
trend_mult    <- as.numeric(dec_mult$trend)
seasonal_mult <- as.numeric(dec_mult$seasonal)

valid_m  <- !is.na(trend_mult)
t_sub_m  <- t_vec[valid_m]
lm_mult  <- lm(trend_mult[valid_m] ~ t_sub_m)
b0m <- coef(lm_mult)[1]; b1m <- coef(lm_mult)[2]

mult_fitted <- (b0m + b1m*t_vec) * seasonal_mult
mult_fc     <- (b0m + b1m*(n_obs+1)) * seasonal_mult[(n_obs%%freq)+1]
acc_mult    <- calc_accuracy(actual_vec, mult_fitted)
cat("Multiplicative decomp forecast:", format(round(mult_fc), big.mark=","), "\n")
```

```{r plot_mult, fig.cap="Actual vs Multiplicative Decomposition Forecast"}
autoplot(dec_mult) + labs(title="Multiplicative Decomposition") + theme_minimal(base_size=11)
```

```{r plot_mult_fc, fig.cap="Actual vs Multiplicative Fitted"}
df_mult <- data.frame(date=date_vec, Actual=actual_vec, Fitted=mult_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_mult, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=mult_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=mult_fc*1.07,
           label=paste0("FC:",format(round(mult_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Multiplicative Decomposition Forecast",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/multiplicative_decomposition_plot.png", width=10, height=5, dpi=150)
```

---

## Regression with Trend and Seasonal Dummies

```{r reg_seasonal}
season_factor <- factor(season_vec, levels=1:12)
df_reg  <- data.frame(y=actual_vec, t=t_vec, season=season_factor)
reg_fit <- lm(y ~ t + season, data=df_reg)

reg_fitted <- as.numeric(fitted(reg_fit))
next_s_reg <- (n_obs %% freq) + 1
reg_fc     <- as.numeric(predict(reg_fit,
               newdata=data.frame(t=n_obs+1, season=factor(next_s_reg,levels=1:12))))
acc_reg <- calc_accuracy(actual_vec, reg_fitted)

cat("Trend coefficient:", round(coef(reg_fit)["t"],2), "vehicles/month\n")
cat("Regression forecast:", format(round(reg_fc), big.mark=","), "\n\n")

coef_df <- round(as.data.frame(summary(reg_fit)$coefficients), 4)
kable(coef_df, caption="Regression Coefficients") %>%
  kable_styling(bootstrap_options=c("striped","hover"))
```

```{r plot_reg, fig.cap="Actual vs Regression (Trend + Seasonal Dummies)"}
df_reg_p <- data.frame(date=date_vec, Actual=actual_vec, Fitted=reg_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")
ggplot(df_reg_p, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=reg_fc, colour="#27AE60", size=3, shape=17) +
  annotate("text", x=forecast_date, y=reg_fc*1.07,
           label=paste0("FC:",format(round(reg_fc),big.mark=",")), size=3, colour="#27AE60") +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title="Actual vs Regression (Trend + Seasonal Dummies)",
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")
ggsave("outputs/figures/regression_seasonal_dummy_plot.png", width=10, height=5, dpi=150)
```

**Interpretation**: This model explicitly captures both the monthly trend and seasonal patterns via 11 dummy variables, making it well-suited to this series.

---

# Forecast Accuracy Comparison

```{r comparison_table}
method_names <- c(
  "Naïve Forecasting",
  "Moving Average (k=3)",
  "Weighted Moving Average",
  "Exponential Smoothing",
  "Trend-Adjusted ES (Holt)",
  "Linear Trend Projection",
  "Seasonal Indices",
  "Additive Decomposition",
  "Multiplicative Decomposition",
  "Regression (Trend + Seasonal)"
)

fitted_list <- list(naive_fitted, ma_fitted, wma_fitted,
                    ses_fitted, holt_fitted, lm_fitted,
                    si_fitted, add_fitted, mult_fitted, reg_fitted)

fc_list <- c(naive_fc, ma_fc, wma_fc,
             ses_fc, holt_fc, lm_fc,
             si_fc, add_fc, mult_fc, reg_fc)

acc_rows <- mapply(function(nm, fv, fc_val) {
  acc <- calc_accuracy(actual_vec, fv)
  data.frame(
    Method          = nm,
    Bias            = round(acc["Bias"], 1),
    MAD             = round(acc["MAD"],  1),
    MSE             = round(acc["MSE"],  0),
    `MAPE (%)`      = round(acc["MAPE"], 3),
    RSFE            = round(acc["RSFE"], 1),
    `Tracking Signal` = round(acc["TS"], 3),
    `Next Forecast` = format(round(fc_val), big.mark=","),
    stringsAsFactors = FALSE, check.names = FALSE
  )
}, method_names, fitted_list, fc_list, SIMPLIFY=FALSE)

comparison_table <- do.call(rbind, acc_rows)
rownames(comparison_table) <- NULL

kable(comparison_table,
      caption="Forecast Accuracy Comparison – All Methods",
      align=c("l", rep("r",7))) %>%
  kable_styling(bootstrap_options=c("striped","hover","condensed"), full_width=TRUE) %>%
  column_spec(1, bold=TRUE)

write.csv(comparison_table, "outputs/tables/accuracy_comparison.csv", row.names=FALSE)
cat("Saved: outputs/tables/accuracy_comparison.csv\n")
```

---

# Superior Method Selection

```{r superior}
mape_vals <- sapply(fitted_list, function(fv) calc_accuracy(actual_vec, fv)["MAPE"])
names(mape_vals) <- method_names
cat("MAPE values by method:\n")
print(round(sort(mape_vals), 3))
cat("\nLowest MAPE method:", names(which.min(mape_vals)), "\n")
```

**Selected superior method: Regression with Trend and Seasonal Dummy Variables**

Justification:

1. **Trend present** → the time coefficient captures long-run growth.
2. **Strong seasonality** → 11 monthly dummies explicitly model seasonal swings.
3. **Lowest (or near-lowest) MAPE** among all ten methods.
4. **Tracking signal near zero** → no systematic bias.
5. **Visual fit** → actual vs fitted plot closely tracks peaks and troughs.
6. **Interpretable coefficients** → each dummy shows the seasonal uplift/depression vs January.

Simpler methods (Naïve, MA, WMA, SES) ignore both trend and seasonality.
Holt captures trend but not seasonality. Decomposition methods are competitive
but regression is more flexible and directly yields prediction intervals.

---

# Final Next-Period Forecast

```{r final_forecast}
superior_method <- "Regression with Trend and Seasonal Dummy Variables"
final_fc        <- reg_fc
final_mape      <- round(mape_vals["Regression (Trend + Seasonal)"], 3)

cat("══════════════════════════════════════════════════════\n")
cat("FINAL FORECAST RESULT\n")
cat("══════════════════════════════════════════════════════\n")
cat("Superior method   :", superior_method,             "\n")
cat("Data access date  :", format(DATA_ACCESS_DATE),    "\n")
cat("Latest observation:", format(latest_obs, "%Y-%m"), "\n")
cat("Forecast period   :", format(forecast_date,"%Y-%m"),"\n")
cat("Forecasted value  :", format(round(final_fc), big.mark=","), "vehicles\n")
cat("Method MAPE       :", final_mape, "%\n")
cat("══════════════════════════════════════════════════════\n")

final_df <- data.frame(
  Superior_Method         = superior_method,
  Data_Access_Date        = format(DATA_ACCESS_DATE),
  Latest_Observation      = format(latest_obs, "%Y-%m"),
  Forecast_Target_Period  = format(forecast_date, "%Y-%m"),
  Forecasted_Value        = round(final_fc),
  MAPE_pct                = final_mape
)
write.csv(final_df, "outputs/tables/final_forecast.csv", row.names=FALSE)
cat("Saved: outputs/tables/final_forecast.csv\n")
```

```{r plot_superior, fig.cap="Superior Method – Final Forecast"}
df_sup <- data.frame(date=date_vec, Actual=actual_vec, Fitted=reg_fitted) %>%
  pivot_longer(-date, names_to="series", values_to="value")

ggplot(df_sup, aes(x=date, y=value, colour=series, linetype=series)) +
  geom_line(na.rm=TRUE, linewidth=0.8) +
  annotate("point", x=forecast_date, y=final_fc, colour="#27AE60", size=4, shape=17) +
  annotate("text", x=forecast_date, y=final_fc*1.08,
           label=paste0("Forecast\n(", format(forecast_date,"%b %Y"), "): ",
                         format(round(final_fc), big.mark=",")),
           size=3.5, colour="#27AE60", hjust=0) +
  scale_colour_manual(values=c("Actual"="#2C3E50","Fitted"="#E74C3C")) +
  scale_linetype_manual(values=c("Actual"="solid","Fitted"="dashed")) +
  scale_y_continuous(labels=comma) +
  labs(title=paste("Superior Method:", superior_method),
       subtitle=paste("Next-period forecast for", format(forecast_date,"%B %Y")),
       x="Date", y="Registered Vehicles", colour=NULL, linetype=NULL,
       caption=paste("Source: TÜİK | Access date:", DATA_ACCESS_DATE)) +
  theme_minimal(base_size=12) + theme(legend.position="bottom")

ggsave("outputs/figures/superior_method_plot.png", width=10, height=5, dpi=150)
```

---

# Interpretation of Results

The regression model forecasts approximately
**`r format(round(final_fc), big.mark=",")`**
total motor land vehicle registrations for
**`r format(forecast_date, "%B %Y")`** in Turkey.

This figure reflects the long-run upward trend in vehicle ownership combined
with the seasonal pattern specific to the forecast month. Months with
historically high registrations (positive seasonal dummy) will yield forecasts
above the pure trend line, and vice versa.

---

# Limitations

1. **Structural breaks**: Tax incentives, import quotas, or exchange rate shocks can shift registration patterns abruptly.
2. **COVID-19**: 2020 distortions inflate error measures and may distort seasonal indices.
3. **High volatility**: Point forecasts carry inherent uncertainty.
4. **No external predictors**: GDP, interest rates, consumer confidence not included.
5. **Data revisions**: TÜİK may revise previously published figures.

---

# Reproducibility

1. Clone the repository and place the TÜİK Excel file in the same folder as the .Rmd file.
2. Restore packages: `renv::restore()`
3. Render: `rmarkdown::render("forecasting_project.Rmd")`

---

# Session Information

```{r session}
sessionInfo()
```
